Visualizations: Metropolitan Areas per Capita, Wave I, II, III (Deaths)

### 2020-11-25 KPMG Team/Dashboard/visualization_scripts/msa-waves-pax.Rmd
msa and region waves I, II, III

# MSA Things
# from Sydney
msa_index <- read_csv("data/msa/msa_mix_index.csv") # csa and msa with population 
covid <- read_csv("data/msa/us_daily_msa_split_2020-11-16.csv") # split

@@

# View(msa)
```

@@

# # left join index of 50 with fips codes
# msa_index_fips <- msa_index %>% 
#   left_join(msa_w_fips, by = c("CSA Title", "County/County Equivalent", "State Name")) %>% 
#   rename(county = `County/County Equivalent`) %>% 
#   rename(csa = `CSA Title`) %>%
#   rename(state = `State Name`) %>%
#   rename(agg_index = mix_index)

msa_index_2 <- msa_index %>% 
  rename(county = `County/County Equivalent`) %>% 
  rename(csa = `CSA Title`) %>%
  rename(state = `State Name`) %>%
  rename(agg_index = mix_index) %>% 
  rename(fips = FIPS_combined) %>% 
  names()

# anti join to get the counties not in msa_index 
region_fips <- msa_w_fips %>% 
  anti_join(msa_index, by = "CSA Title") %>% 
  rename(county = `County/County Equivalent`) %>%
  mutate(agg_index = `State Name`) %>%
  rename(state = `State Name`) %>%
  rename(csa = `CSA Title`)

# joining fips info with covid dataset (has states + selected msas)
state_msa_fips_join <- covid %>% 
  full_join(msa_index_fips, by = "agg_index")# %>%

@@

# keeping only states, don't have counties
state_covid_only <- state_msa_fips_join %>% 
  anti_join(msa_covid_only) %>% 
  select(-fips, -mix_index_population, -FIPS_population, -FIPS_combined, -`Central/Outlying County`, -state, -county, -csa, -`Metropolitan Division Title`)

names(msa_region)

region_fips_join <- msa_region %>% 
  full_join(region_fips, by = "agg_index") %>%
  drop_na(fips)

View(msa_covid_only)

View(region_fips_join)
View(covid)

final_daily <- msa_fips_join %>% # or final total 
  bind_rows(region_fips_join) %>% 
  select(index, Date, state, agg_index, fips, csa, county, population, 
         Confirmed, Deaths, new_confirmed, new_deaths, 
         new_confirmed_7, new_deaths_7, new_confirmed_14, new_deaths_14,
         per_100k_new_confirmed_7, per_100k_new_deaths_7)

names(final_daily)
# write.csv(final_daily, "data/final-50-county-daily.csv")

@@

# add lat long attributes to county dataset          
county_geo <- merge(county_lines, final_daily, by = "fips")
county_map <- county_geo %>% st_transform('+proj=longlat +datum=WGS84')

map_daily <- county_map # daily covid map

# Wave/Dates Filter  
### For total cases

```{r}
# library(spdplyr) # spatial version of dplyr

# Dates of Waves

wave_2 <- as.Date("2020-06-15") # start of wave 2
wave_3 <- as.Date("2020-10-15") # start of wave 3

# all cases map 
map_total <- county_map %>% 
  filter(Date == max(as.Date(county_map$Date), na.rm = TRUE)) %>% # need to remove NA
  rename(state_fips = STATEFP) %>%
  select(-new_confirmed, -new_deaths, -new_confirmed_7, -new_deaths_7, -new_confirmed_14, -new_deaths_14, -per_100k_new_confirmed_7, -per_100k_new_deaths_7, -COUNTYNS, -AFFGEOID, -GEOID, -county)

# wave 2 start date 
map_w2_start <- map_daily %>%
  filter(Date == "2020-06-15") %>% 
  rename(state_fips = STATEFP) %>%
  select(-new_confirmed, -new_deaths, -new_confirmed_7, -new_deaths_7, -new_confirmed_14, -new_deaths_14, -per_100k_new_confirmed_7, -per_100k_new_deaths_7, -COUNTYNS, -AFFGEOID, -GEOID, -county)

# wave 2 map 
map_w2 <- map_total %>% 
  st_join(map_w2_start, by = "fips") %>% 
  mutate(Confirmed = Confirmed.x - Confirmed.y) %>% 
  mutate(Deaths = Deaths.x - Deaths.y) %>% 
  rename(start_date = Date.y) %>% 
  rename(end_date = Date.x) %>% 
  rename(fips = fips.x) %>% 
  rename(county = NAME.x) %>% 
  rename(population = population.x) %>% 
  rename(csa = csa.x) %>% 
  rename(agg_index = agg_index.x) %>% 
  rename(state = state.x) %>%
  rename(index = index.x) %>%
  rename(county_fips = COUNTYFP.x) %>%
  rename(state_fips = state_fips.x) %>%
  select(fips, index, agg_index, state, county, csa, population, Confirmed, Deaths, start_date, end_date, state_fips, county_fips, geometry, LSAD.x, ALAND.x, AWATER.x)

@@

```{r}
library(usmap) # to get US Map Data

names(map_w2)
fips("New York")
#state_name,

# county state map 
mapping <- function(map_type, pax) {
  # 
  # fips_code <- fips(state = state_name) 
  # 
  # map <- map_type %>% 
  #   filter(state_fips == fips_code) %>% 
  #   filter(Deaths >= 0)
  #

  map <- map_type

  ggplot(data = map) +
    geom_sf(aes(fill = Deaths, color = Deaths)) + 
    scale_fill_viridis_c(option = "viridis", direction = -1) + 
    labs(title = "something", #paste(state),
         subtitle = "Total Deaths") +
    theme_void() + 
    theme(text = element_text(family = "raleway"))

  # if (pax = FALSE) {
  #   
  #   map <- 
  #   
  # } else if (pax = TRUE) {
  #   
  #   
  #

}

dim(map_w2)

length(map_total$Deaths)
length(map_w2$Deaths)

mapping(map_w2)

@@

>>>>

# MSA Things
# from Sydney
msa_index <- read_csv("data/msa/msa_mix_index.csv") # csa and msa with population 
covid <- read_csv("data/msa/us_daily_msa_split_2020-11-16.csv") # split 
covid_state <- read_csv("data/us_state_level_clean_2020-11-20.csv") # states only

@@

# View(msa)

covid_state
```

@@

msa_index_2 <- msa_index %>% 
  rename(county = `County/County Equivalent`) %>% 
  rename(csa = `CSA Title`) %>%
  rename(state = `State Name`) %>%
  rename(agg_index = mix_index) %>% 
  rename(fips = FIPS_combined)

# joining fips info with covid dataset (has states + selected msas)
state_msa_fips_join <- covid %>% 
  full_join(msa_index_2, by = "agg_index")# %>%

@@

# keeping only states, don't have counties
state_covid_only <- state_msa_fips_join %>% 
  anti_join(msa_covid_only) %>%
  select(-fips, -mix_index_population, -FIPS_population, -`Central/Outlying County`, -state, -county, -csa, -`Metropolitan Division Title`)

# Getting Wave Count

## for States

```{r}
wave_2 <- as.Date("2020-06-15") # start of wave 2
wave_3 <- as.Date("2020-10-15") # start of wave 3

state_covid_reg <- covid_state %>% 
  rename("NAME" = "Province_State") %>%
  filter(! NAME %in% c("Alaska", "Hawaii", "Rhode Island", "District of Columbia"))

state_reg_total <- state_covid_reg %>% 
  filter(Date == max(as.Date(state_covid_reg$Date), na.rm = TRUE)) %>% # need to remove NA
  mutate(Date_total = Date)

state_reg_w2_start <- state_covid_reg %>% 
  filter(Date == wave_2) %>%
  mutate(Date_W2 = Date)

state_reg_w3_start <- state_covid_reg %>% 
  filter(Date == wave_3) %>% 
  mutate(Date_W3 = Date)

View(state_reg_w2_start)
View(state_reg_total)
View(state_all)

state_reg_all <- state_reg_total %>% 
  rename(Confirmed_total = Confirmed) %>% # Start - Latest
  rename(Deaths_total = Deaths) %>%
  # joining wave 2 and beyond numbers
  left_join(state_reg_w2_start, by = "NAME") %>%  # Start - June 15
  # wave 1 only %>% 
  rename(Confirmed_W1 = Confirmed) %>%  # Start - June 15
  rename(Deaths_W1 = Deaths) %>% # Start - June 15
  # wave 2 and beyond column 
  mutate(Confirmed_W2_plus = Confirmed_total - Confirmed_W1) %>% # June 15 - Latest
  mutate(Deaths_W2_plus = Deaths_total - Deaths_W1) %>% # June 15 - Latest
  # joining wave 3 numbers 
  left_join(state_reg_w3_start, by = "NAME") %>% # Start - Oct 15 ie Wave 1 & Wave 2
  # only wave 2 
  mutate(Deaths_W2_only = Deaths - Deaths_W1) %>% # June 15 - Oct 15
  mutate(Confirmed_W2_only = Confirmed - Confirmed_W1) %>% # June 15 - Oct 15
  # wave 3 and beyond
  mutate(Deaths_W3 = Deaths_total - Deaths) %>%  # Oct 15 - Latest
  mutate(Confirmed_W3 = Confirmed_total - Confirmed) %>% # Oct 15 - Latest
  # tidy up, selecting only relevant variables
  select(NAME, population, 
         Date_total, Confirmed_total, Deaths_total, 
         Confirmed_W1, Deaths_W1,
         Date_W2, Confirmed_W2_plus, Confirmed_W2_only, Deaths_W2_plus, Deaths_W2_only, 
         Date_W3, Confirmed_W3, Deaths_W3, stringency, containment, RegionCode)

## for States without MSA

state_covid_only_2 <- state_covid_only %>% 
  rename("NAME" = "agg_index") %>%
  filter(! NAME %in% c("Alaska", "Hawaii", "Rhode Island", "District of Columbia"))

state_total <- state_covid_only_2 %>% 
  filter(Date == max(as.Date(state_covid_only_2$Date), na.rm = TRUE)) %>% # need to remove NA
  mutate(Date_total = Date)

state_w2_start <- state_covid_only_2 %>% 
  filter(Date == max(wave_2)) %>%
  mutate(Date_W2 = Date)

state_w3_start <- state_covid_only_2 %>% 
  filter(Date == max(wave_3)) %>% 
  mutate(Date_W3 = Date)

state_all <- state_total %>% 
  rename(Confirmed_total = Confirmed) %>% # Start to Latest
  rename(Deaths_total = Deaths) %>%
  # joining wave 2 and beyond numbers
  left_join(state_w2_start, by = "NAME") %>%  # Start to June 15
  # wave 1 only %>% 
  rename(Confirmed_W1 = Confirmed) %>%  # Start to June 15
  rename(Deaths_W1 = Deaths) %>% # Start to June 15
  # wave 2 and beyond column 
  mutate(Confirmed_W2_plus = Confirmed_total - Confirmed_W1) %>% # June 15 to Latest
  mutate(Deaths_W2_plus = Deaths_total - Deaths_W1) %>% # June 15 to Latest
  # joining wave 3 numbers 
  left_join(state_w3_start, by = "NAME") %>% # Start to Oct 15 ie Wave 1 & Wave 2
  # only wave 2 
  mutate(Confirmed_W2_only = Confirmed - Confirmed_W1) %>% # June 15 to Oct 15
  mutate(Deaths_W2_only = Deaths - Deaths_W1) %>% # June 15 to Oct 15
  # wave 3 and beyond
  mutate(Deaths_W3 = Deaths_total - Deaths) %>%  # Oct 15 to Latest
  mutate(Confirmed_W3 = Confirmed_total - Confirmed) %>% # Oct 15 to Latest
  # tidy up, selecting only relevant variables
  select(index, NAME, population, 
         Date_total, Confirmed_total, Deaths_total, 
         Confirmed_W1, Deaths_W1,
         Date_W2, Confirmed_W2_plus, Confirmed_W2_only, Deaths_W2_plus, Deaths_W2_only, 
         Date_W3, Confirmed_W3, Deaths_W3)

View(msa_all)

## for MSAs/Top 50

msa_covid_only_2 <- msa_covid_only %>% 
  rename("NAME" = "agg_index")

msa_total <- msa_covid_only_2 %>% 
  filter(Date == max(as.Date(msa_covid_only_2$Date), na.rm = TRUE)) %>% # need to remove NA
  mutate(Date_total = Date)

msa_w2_start <- msa_covid_only_2 %>% 
  filter(Date == max(wave_2)) %>%
  mutate(Date_W2 = Date)

msa_w3_start <- msa_covid_only_2 %>% 
  filter(Date == max(wave_3)) %>% 
  mutate(Date_W3 = Date)

msa_all <- msa_total %>% 
  rename(Confirmed_total = Confirmed) %>% # Start - Latest
  rename(Deaths_total = Deaths) %>%
  # joining wave 2 and beyond numbers
  left_join(msa_w2_start, by = "NAME") %>%  # Start - June 15
  # wave 1 only %>% 
  rename(Confirmed_W1 = Confirmed) %>%  # Start - June 15
  rename(Deaths_W1 = Deaths) %>% # Start - June 15
  # wave 2 and beyond column 
  mutate(Confirmed_W2_plus = Confirmed_total - Confirmed_W1) %>% # June 15 - Latest
  mutate(Deaths_W2_plus = Deaths_total - Deaths_W1) %>% # June 15 - Latest
 left_join(msa_w3_start, by = "NAME") %>% # Start - Oct 15 ie Wave 1 & Wave 2
  # only wave 2 
  mutate(Deaths_W2_only = Deaths - Deaths_W1) %>% # June 15 - Oct 15
  mutate(Confirmed_W2_only = Confirmed - Confirmed_W1) %>% # June 15 - Oct 15
  # wave 3 and beyond
  mutate(Deaths_W3 = Deaths_total - Deaths) %>%  # Oct 15 - Latest
  mutate(Confirmed_W3 = Confirmed_total - Confirmed) %>% # Oct 15 - Latest
  # tidy up, rename
  rename(metro_title = "Metropolitan Division Title") %>%
  rename(central_outlying = "Central/Outlying County") %>%
  # selecting only relevant variables
  select(index, NAME, fips, csa, county, metro_title, population, FIPS_population, mix_index_population,
         Date_total, Confirmed_total, Deaths_total,
         Confirmed_W1, Deaths_W1,
         Date_W2, Confirmed_W2_plus, Confirmed_W2_only, Deaths_W2_plus, Deaths_W2_only,
         Date_W3, Confirmed_W3, Deaths_W3) %>%
  distinct()

@@

# add lat long attributes to county dataset          
county_geo <- merge(county_lines, msa_all, by = "fips")
county_map <- county_geo %>% st_transform('+proj=longlat +datum=WGS84')

# merge state 
# add lat long attributes to state dataset

state_geo <- merge(state_lines, state_all, by = "NAME")
state_map <- state_geo %>% st_transform('+proj=longlat +datum=WGS84')

state_reg_geo <- merge(state_lines, state_reg_all, by = "NAME")
state_reg_map <- state_reg_geo %>% st_transform('+proj=longlat +datum=WGS84')

@@

## state

options(scipen = 999)

library(usmap) # to get US Map Data
library(cowplot) # for grid plot

# fips("New York")
#state_name,

g_total <- ggplot() +
  geom_sf(data = state_map, 
          aes(fill = Deaths_total/population), 
          color = NA) +  # color = NA 
  geom_sf(data = county_map, aes(fill = Deaths_total/population), color = NA) + # MSA Things
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1) + 
  labs(title = "Total Deaths\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 50))

max_W1 <- max(c(state_map$Deaths_W1 / state_map$population, county_map$Deaths_W1 / county_map$population))
max_W2 <- max(c(state_map$Deaths_W2_only / state_map$population, county_map$Deaths_W2_only / county_map$population))
max_W3 <- max(c(state_map$Deaths_W3 / state_map$population, county_map$Deaths_W3 / county_map$population))

min_W1 <- min(c(state_map$Deaths_W1 / state_map$population, county_map$Deaths_W1 / county_map$population))
min_W2 <- min(c(state_map$Deaths_W2_only / state_map$population, county_map$Deaths_W2_only / county_map$population))
min_W3 <- min(c(state_map$Deaths_W3 / state_map$population, county_map$Deaths_W3 / county_map$population))

max_W1 
max_W2
max_W3
min_W1
min_W2
min_W3

max(state_map$Deaths_W1 / state_map$population)
max(county_map$Deaths_W1 / county_map$population)
min(state_map$Deaths_W1 / state_map$population)
min(county_map$Deaths_W1 / county_map$population)

max(state_map$Deaths_W2_only / state_map$population)
max(county_map$Deaths_W2_only / county_map$population)
min(state_map$Deaths_W2_only / state_map$population)
min(county_map$Deaths_W2_only / county_map$population)

max(state_map$Deaths_W3 / state_map$population)
max(county_map$Deaths_W3 / county_map$population)
min(state_map$Deaths_W3 / state_map$population)
min(county_map$Deaths_W3 / county_map$population)

g_w1 <- ggplot(data = state_reg_map) +
  geom_sf(aes(fill = Deaths_W1/population), color = NA) + 
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1, limits = c(min_W1, max_W1)) + 
  labs(title = "U.S. States - Wave 1 Deaths per Capita",
      subtitle = "Until 15 June 2020\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 30))

g_w2 <- ggplot(data = state_reg_map) +
  geom_sf(aes(fill = Deaths_W2_only/population), color = NA) + 
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1, limits = c(min_W2, max_W2)) + 
  labs(title = "U.S. States - Wave 2 Deaths per Capita",
       subtitle = "From 15 June 2020 - 15 October 2020\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 30))

g_w3 <- ggplot(data = state_reg_map) +
  geom_sf(aes(fill = Deaths_W3/population), color = NA) + 
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1, limits = c(min_W3, round(max_W3))) + 
  labs(title = "U.S. States - Wave 3 Onwards Deaths per Capita",
       subtitle = "After 15 October 2020\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 30))

# state and county msa 
g_c_w1 <- ggplot() +
  geom_sf(data = state_map, aes(fill = Deaths_W1/population), color = NA) +
  geom_sf(data = county_map, aes(fill = Deaths_W1/population), color = NA) + # MSA Things
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1, limits = c(min_W1, max_W1)) + 
  labs(title = "Metropolitan State Areas, Regional Areas - Wave 1 Deaths per Capita", 
       subtitle = "Until 15 June 2020\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 30))

g_c_w2 <- ggplot(data = state_map) +
  geom_sf(aes(fill = Deaths_W2_only/population), color = NA) + # size = 0.3
  geom_sf(data = county_map, aes(fill = Deaths_W2_only/population), color = NA) + # MSA Things
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1, limits = c(min_W2, max_W2)) +
  labs(title = "Metropolitan State Areas, Regional Areas - Wave 2 Deaths per Capita",
       subtitle = "From 15 June 2020 - 15 October 2020\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 30))

g_c_w3 <- ggplot(data = state_map) +
  geom_sf(aes(fill = Deaths_W3/population), color = NA) + 
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1, limits = c(min_W3, round(max_W3))) + 
    geom_sf(data = county_map, aes(fill = Deaths_W3/population), color = NA) + # MSA Things
  labs(title = "Metropolitan State Areas, Regional Areas - Wave 3 Deaths per Capita",
       subtitle = "After 15 October 2020\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 30))

# Grid Plots 
p <- plot_grid(g_total, g_w2, g_w3, ncol = 3)

p_2 <- plot_grid(g_w2, g_c_w2, ncol = 2)
p_3 <- plot_grid(g_w3, g_c_w3, ncol = 2)

title <- ggdraw() + 
  draw_label("Total Deaths per Capita in U.S. by State per Wave", fontface='bold') + 
  theme(text = element_text(family = "raleway"))

plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1))

# ggsave("viz/state-waves.png", width = 8, height = 3, dpi = 300)
# ggsave("viz/state-waves-total.png", g_total, width = 8, height = 5, dpi = 300)
# ggsave("viz/state-waves-W1.png", g_w1, width = 8, height = 5, dpi = 300)
# ggsave("viz/state-waves-W2.png", g_w2, width = 8, height = 5, dpi = 300)
# ggsave("viz/state-waves-W3.png", g_w3, width = 8, height = 5, dpi = 300)

ggsave("viz/state-wave-w1.png", g_w1, width = 8, height = 5, dpi = 300)
ggsave("viz/state-msa-wave-w1.png", g_c_w1, width = 8, height = 5, dpi = 300)

ggsave("viz/state-wave-w2.png", g_w2, width = 8, height = 5, dpi = 300)
ggsave("viz/state-msa-wave-w2.png", g_c_w2, width = 8, height = 5, dpi = 300)

ggsave("viz/state-wave-w3.png", g_w3, width = 8, height = 5, dpi = 300)
ggsave("viz/state-msa-wave-w3.png", g_c_w3, width = 8, height = 5, dpi = 300)

ggsave("viz/state-msa-waves-W1.png", plot_grid(g_w1, g_c_w1, ncol = 2), width = 14, height = 5, dpi = 300)
ggsave("viz/state-msa-waves-W2.png", plot_grid(g_w2, g_c_w2, ncol = 2), width = 14, height = 5, dpi = 300)
ggsave("viz/state-msa-waves-W3.png", plot_grid(g_w3, g_c_w3, ncol = 2), width = 14, height = 5, dpi = 300)

# ggsave("viz/state-msa-waves-temp.png", width = 14, height = 5, dpi = 300)

@@

## top 50

c_total <- ggplot(data = county_map) +
  geom_sf(aes(fill = Deaths_total/population), size = 0.3) +  # color = NA 
  scale_fill_viridis_c(name = "Deaths per Capita", option = "viridis", direction = -1) + 
  labs(title = "Total Deaths\n") + 
  theme_void() + 
  theme(text = element_text(family = "raleway", size = 50))
